Fixing Vulnerabilities Hinders Maintainability (REG)

### 2019-05-09 paper/ICPC19/icpc19.tex
paper issues addressed

\documentclass[10pt,conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{xfrac}

@@

\title{Fixing Vulnerabilities Hinders Maintainability}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their codebases, they actually
affect the maintainability of software, potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
dataset including 607 security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
need to be performed with extra care.
\end{abstract}

@@

Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications. Software
quality characteristics include, but are not limited to, functional correctness,
reliability, usability, maintainability, and security. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate
systems and exploit their resources.
As an example, in the last week of January $2019$, a severe vulnerability in Apple
FaceTime was revealed\footnote{FaceTime vulnerability details available at \url{https://thehackernews.com/2019/01/apple-facetime-
privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
flaw that allowed callers to see and hear others without them picking up the call.

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics.
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
prevents the introduction of vulnerabilities in the codebases. In this work, we
are concerned about the impact that patching vulnerabilities may have in the
open-source software maintainability. Are developers affecting software
maintainability when trying to improve software security? We suspect that some
patches may have a negative impact on the software maintainability and
possibly even be the cause of the introduction of new vulnerabilities.

ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. As ISO does not provide any specific guidelines
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
project. BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model).

The particular subject of this paper is to explore whether there is a trade-off
between refactoring source code to patch disclosed vulnerabilities and keeping
(or even improving) the maintainability of software systems. Hedegus et al.
($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
groups of refactored elements and non-refactored and conclude that the source
code elements subjected to refactorings had significantly lower maintainability
than the elements not modified.

In this paper, we present the results of our analysis on the maintainability of
$607$ security changes collected from open-source software, from different
programming languages. The main contributions of the present work are:

@@

This empirical study exhibits proof that changes applied in the codebase to fix
vulnerabilities affect code maintainability. Especially, for refactorings that
address \emph{Broken Authentication and Session Management} issues, \emph{Memory
Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight
the need for tools to predict the impact of patches on maintainability~\cite{4724577}
and documentation to help developers adopt security best practices~\cite{6311252,
7927935, MESQUIDA201519}.

This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and section~\ref{sec:discussion} discusses their
implications; section~\ref{sec:threats}, enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security refactorings impact
the maintainability of software.

@@

  static int dtls1_reassemble_fragment(SSL *s, struct hm_header_st* msg_hdr, int *ok){

@@

- if((msg_hdr->frag_off+frag_len) > max_len)

+ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
+ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
     goto err; /*!\annotation{lst:func5}!*/

  memset(seq64be,0,sizeof(seq64be));
  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
  seq64be[7] = (unsigned char) msg_hdr->seq;
  item = pqueue_find(s->d1->buffered_messages, seq64be);
  if(item == NULL)
  {
    frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);

@@

 static int
 dtls1_process_out_of_seq_message(SSL *s, struct hm_header_st* msg_hdr, int *ok)
 {

@@

  if(frag_len && frag_len < msg_hdr->msg_len)
     return dtls1_reassemble_fragment(s, msg_hdr, ok);

@@

Although checking the maximum size of a handshake message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the codebase which produced a negative impact in the
maintainability. Even though the developer tries to simplify the
codebase by introducing a smaller unit
\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
contributes to a potential violation of the \emph{Keep Your Codebase Small}
guideline. Hence, we have observed that the impact of this refactoring on the
OpenSSL maintainability was negative.

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security refactorings on the
maintainability of open-source software?}
\end{framed}

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the codebase to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $607$ commits using the
\emph{Better Code Hub} tool. The same approach is applied to a randomly generated
dataset of regular commits --- baseline --- that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
affect open-source software maintainability?}
\end{framed}
% here
There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a
cross-site scripting vulnerability since usually the last one can be fixed
without adding new lines of code. A typical fix for the cross-site scripting
vulnerability is presented in Listing~\ref{lst:fix}. The developer added the
function \texttt{htmlentities} to \textit{escape} the data given by the variable
\texttt{\$\_['file']}. Knowing which patterns are more likely to increase
maintainability issues is one step forward to bring awareness to security
engineers of what patterns need more attention. The taxonomy of security
patterns used to answer this question is the same as the one provided by the
authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
separately for each pattern.

@@

\begin{enumerate}
	\item Extract relevant data from a dataset containing $711$ security
	refactorings collected from open-source software available on
	GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was randomly collected from the list of
	projects of the main dataset to evaluate the impact of regular commits on the
	maintainability of open-source software.
  \item Use the Software Improvement Group's\footnote{SIG's website is available
  at \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
  service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
  \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
  for both security and regular commits.
\end{enumerate}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.4\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu ($2017$) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset with $711$ test cases for $16$ security patterns. Some of the patterns
are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
$2017$~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
 	\caption{Distribution of Security Refactorings per Language}
	\label{fig:lang}
\end{figure}

@@

\begin{table*}[h]
	\centering
	\caption{Descriptive statistics of the projects in the dataset}
\begin{tabular}{@{}rrrrrrrrrrr@{}}
\toprule
      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
\end{tabular}
\label{tab:dataset}
\end{table*}

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintainability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. These thresholds are calibrated by BCH using their own
data/experience --- using open-source and closed software systems. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
\url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.

@@

\begin{table}[h]
	\caption{Guidelines to produce maintainable code}
\begin{tabular}{L{2cm}L{4cm}L{1cm}}

\toprule
\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
\midrule
\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
 units are easier to understand, reuse and test them & \\\midrule
\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
it makes units easier to test and modify & \\\midrule
\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
multiple places (inefficient and error-prone) &\\\midrule
\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
$4$ because it makes units easier to understand and reuse & \\\midrule
\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
loosely coupled databases are easier to oversee and execute & \\\midrule
\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
within modules that are exposed to modules in other components & \\\midrule
\textbf{Keep Architecture Components Balanced} & Balancing the number of
components ease locating code and allow for isolated maintenance & \\\midrule
\textbf{Keep your Codebase Small} & Reduce and avoid the system size because
small products are easier to manage and maintain & \\\midrule
\textbf{Automate Tests} & Test your codebase because it makes development
predictable and less risky & \\\midrule
\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future & \\
\bottomrule
\end{tabular}
\label{tab:guidelines}
\end{table}

@@

\begin{equation}
    w(l) = \frac{1 - T(l)}{T(l)}
\end{equation}

\noindent
where $T(l)$ is the threshold in percentage of the lines of code that are
accepted to be non-compliant with the guideline for the severity level $l$. This
is a standard value defined by BCH. In other words, the factor $w$ is used in
Equation~\ref{eq:3} to highlight the lines of code that are not complying with
the guideline. Then, we compute the difference of maintainability between the
security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
Figure~\ref{fig:commit}.

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security refactorings on the
maintainability of open-source software?}
\end{framed}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability.pdf}
 	\caption{Maintainability difference for security and baseline refactorings}
	\label{fig:secvsreg}
\end{figure}

@@

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
affect open-source software maintainability?}
\end{framed}

>>>>

\documentclass[10pt,conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}
\usepackage{framed}
\usepackage{float}
\usepackage{xcolor}
\usepackage{footnote}
\usepackage[hidelinks]{hyperref}
\usepackage{xfrac}

@@

\title{Fixing Vulnerabilities Hinders Maintainability (REG)}

@@

\begin{abstract}
Security is a requirement of utmost importance to produce software with high
quality. However, there is still a considerable amount of zero-day
vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
developers address these vulnerabilities on their code bases, they actually
affect the maintainability and, consequently, the reliability of software $-$ potentially introducing other
vulnerabilities. This paper evaluates the impact of refactorings to improve
security on the maintainability of open-source software. Maintainability is
measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
dataset including 607 security commits. Results show that fixing software
vulnerabilities hinders the maintainability of open-source software. This is
particularly the case for refactorings that address \emph{Broken Authentication
and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
attacks. Furthermore, we conclude that changes to code bases to fix vulnerabilities
need to be performed with extra care in open-source software.
\end{abstract}

@@

Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. Software
quality characteristics include, but are not limited to functional correctnes,
reliability, usability, maintainability, and security~\cite{iso:2011}. Security is an important
non-functional requirement during the development of software systems.  However,
there is still a considerable amount of zero-day vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate
systems and exploit their resources.
As an example, in the last week of January $2019$, a severe vulnerability in Apple
FaceTime was revealed\footnote{FaceTime vulnerability details available at \url{https://thehackernews.com/2019/01/apple-facetime-
privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
flaw that allowed callers to see and hear others without them picking up the call.

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
\emph{Security} as one of the main software product quality characteristics. Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
supports the production of more reliable software preventing the introduction of faults/vulnerabilities in the code bases\Sof{cite smth}. ISO describes software maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010. Software maintainbility can also be defined as the property of
software that reveals how easily the software system can be maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.

There is still little information about the impact of patching vulnerabilities
in open-source software maintainability. 
% Hedegus et al.
% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
% groups of refactored elements and non-refactored and conclude that the source
% code elements subjected to refactorings had significantly lower maintainability
% than the elements not modified.
\textit{Are developers affecting software
maintainability when trying to improve software security?}, it is what we want to investigate. Our suspiciousness is
that some of the patches may have a negative impact on the software maintainability and possibly even be the cause of the introduction
of new vulnerabilities $-$ harming software reliability. Therefore, in this paper we present an empirical study on the evolution of patches of vulnerabilities accross open-source software.

As ISO does not provide any specific guidelines
or formula to calculate maintainability, we resort on Software Improvement
Group (SIG)'s web-based source code analysis service Better Code Hub
(BCH) to compute the software compliance with a set of $10$ guidelines to
produce quality software~\cite{Visser:2016:OREILLY} based in ISO/IEC 25010. The results obtained for each guideline are merged into a mathematical formula that calculates software maintainability. From a methodological perspective, we leveraged a dataset of $607$ security changes collected from open-source software. Each change provides two different versions of the software - the vulnerable one and the patched one. For each version, we calculate software maintainability, using the proposed formula, after BCH computes the compliance with each guideline. Thus, we assess the impact of security changes comparing the maintainbility value of both versions. We also compare the impact of security changes with regular changes on software
maintainability. For that, we also collect a dataset of regular changes and do the same experiments as for security changes. This empirical study exhibits proof that changes applied in the code bases to fix
vulnerabilities affect code maintainability. Especially, for refactorings that
address \emph{Broken Authentication and Session Management} issues, \emph{Memory
Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight
the need for tools to predict the impact of patches on maintainability~\cite{4724577}
and documentation to help developers adopt security best practices~\cite{6311252,
7927935, MESQUIDA201519}.

%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.

% There are other well-known standards and models that have been proposed
% to increase software security: Common Criteria~\cite{common:2009} which received
% negative criticism regarding the costs associated and poor technical evaluation;
% the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
% focused only on web applications, and a model proposed by Xu et al. ($2013$)
% ~\cite{6616351} for rating software security (arguably, it was one of the
% first steps taken by SIG to introduce security on their maintainability model).

The main contributions of the present work are:

@@

This paper is structured as follows: section~\ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at \url{https://github.com/openssl/openssl}
(Accessed on \today{})}; section~\ref{sec:methodology} describes the
methodology used to answer the research questions; section~\ref{sec:results}
presents the results and section~\ref{sec:discussion} discusses their
implications; section~\ref{sec:threats}, enumerates the threats to the validity of
this study; section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, section~\ref{sec:conclusions}
concludes the main findings and elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software code base. However, these refactorings may have a
negative impact on software maintenance --- e.g., developers might follow the
quickest solution to fix the problem and not necessarily the most elegant and
performant one. Thus, we want to understand how security refactorings impact
the maintainability of software.

@@

  static int dtls1_reassemble_fragment(
     SSL *s, struct hm_header_st* msg_hdr, int *ok){

@@

- if((msg_hdr->frag_off+frag_len) > max_len)
+ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
+ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
     goto err; /*!\annotation{lst:func5}!*/

  memset(seq64be,0,sizeof(seq64be));
  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
  seq64be[7] = (unsigned char) msg_hdr->seq;
  item = pqueue_find(s->d1->buffered_messages, seq64be);
  if(item == NULL){
     frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);

@@

 static int dtls1_process_out_of_seq_message(
     SSL *s, struct hm_header_st* msg_hdr, int *ok){

@@

@@

Although checking the maximum size of a handshake message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the code base which produced a negative impact in the
maintainability. Even though the developer tries to simplify the
code base by introducing a smaller unit
\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. Moreover, it increases the code base
size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
contributes to a potential violation of the \emph{Keep Your code base Small}
guideline. Hence, we have observed that the impact of this refactoring on the
OpenSSL maintainability was negative.

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security refactorings on the
maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.2cm}

Often security flaws require refactoring code to make software more secure.
However, there is no evidence yet of how security refactorings impact the
maintainability of open-source software. Our suspicion is that developers tend
to introduce technical debt in their software when refactoring the code base to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $607$ commits using the
\emph{Better Code Hub} tool. The same approach is applied to a randomly generated
dataset of regular commits --- baseline --- that we use to understand how
maintainability evolves when security refactorings are performed versus when
they are not.

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
affect open-source software maintainability?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.2cm}
% here
There are security flaws that are more difficult to refactor than others. For
instance, implementing secure authentication is not as easy as fixing a
cross-site scripting vulnerability since usually the last one can be fixed
without adding new lines of code. A typical fix for the cross-site scripting
vulnerability is presented in Listing~\ref{lst:fix}. The developer added the
function \texttt{htmlentities} to \textit{escape} the data given by the variable
\texttt{\$\_['file']}. Knowing which patterns are more likely to increase
maintainability issues is one step forward to bring awareness to security
engineers of what patterns need more attention. The taxonomy of security
patterns used to answer this question is the same as the one provided by the
authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
separately for each pattern.

@@

\begin{enumerate}
	\item Extract relevant data (e.g., sha, owner of the project, name of the repository) and code bases from a dataset containing $711$ security
	refactorings collected from open-source software available on
	GitHub~\cite{Reis:2017:IJSSE}.
	\item A baseline of regular commits was randomly collected from the list of
	projects of the main dataset to evaluate the impact of regular commits on the
	maintainability of open-source software.
  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
  service \emph{Better Code Hub} (BCH)\footnote{BCH's website:
  \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
  for both security and regular commits. The code bases that resemble to the commits
  of our datasets are collected before the analysis performed by BCH.

\end{enumerate}
\vspace{-0.3cm}

%
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.4\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}

\end{figure}
	\vspace{-0.2cm}

%
\subsection{Dataset}
%
To evaluate the impact of security refactorings on the maintainability of
open-source software, we use a dataset of security flaws which is the outcome of
mining $94$ GitHub projects. Reis and Abreu ($2017$) mined open-source
software aiming at the extraction of real examples --- created by real
developers --- to test and assess the performance of static analysis tools since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset with $711$ test cases for $16$ security patterns. Some of the patterns
are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
$2017$~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the refactoring, the commit responsible
for the refactoring, and the snippets of code that differ from one version to
another (typically, called \textit{diffs}) --- where one can easily review the
code used to fix the security flaw. In this study, we focus on computing the
maintainability of the commits before and after the security refactoring to
evaluate if the impact was positive, negative or none.
\vspace{-0.2cm}

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
 	\caption{Distribution of Security Refactorings per Language}
	\label{fig:lang}
\end{figure}
\vspace{-0.2cm}

@@

\begin{table*}[h]
	\centering
	\caption{Descriptive statistics of the projects in the dataset}
\begin{tabular}{@{}rrrrrrrrrrr@{}}
\toprule
      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
\end{tabular}
\vspace{-0.3cm}
\label{tab:dataset}
\end{table*}

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintainability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines $-$ proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric calculated by BCH. Some of theses guidelines are calculated using the metrics presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. These thresholds are calibrated by BCH using their own
data/experience --- using open-source and closed software systems. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
\url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.

@@

\begin{table}[h]
	\caption{Guidelines to produce maintainable code}
\begin{tabular}{L{1.66cm}L{3.25cm}L{2.65cm}}

\toprule
\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
\midrule

\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
 units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of LOCs within each unit~\cite{criteria:2017} \\\midrule

\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision points~\cite{1702388,criteria:2017}\\\midrule

\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant LOCs~\cite{criteria:2017}\\\midrule

\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} \# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule

\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
incoming dependencies~\cite{criteria:2017} \\\midrule

\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
within modules that are exposed to modules in other components & \textbf{Component Independence:} \% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule

\textbf{Keep Architecture Components Balanced} & Balancing the number of
components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule

\textbf{Keep your code base Small} & Reduce and avoid the system size because
small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted to man-month/man-year~\cite{criteria:2017} \\\midrule

\textbf{Automate Tests} & Test your code base because it makes development
predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit complexity, component independence and volume~\cite{Visser:2016:OREILLY}
 \\\midrule

\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future & Help ? \\
\bottomrule
\end{tabular}
\vspace{-0.5cm}
\label{tab:guidelines}
\end{table}

@@

\begin{equation}
    w(l) = \frac{1 - \theta(l)}{\theta(l)}
\end{equation}

\noindent
where $\theta(l)$ is the threshold in percentage of the lines of code that are
accepted to be non-compliant with the guideline for the severity level $l$. This
is a standard value defined by BCH. In other words, the factor $w$ is used in
Equation~\ref{eq:3} to highlight the lines of code that are not complying with
the guideline. Then, we compute the difference of maintainability between the
security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
Figure~\ref{fig:commit}.

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security refactorings on the
maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.2cm}

@@

\vspace{-0.3cm}
\begin{figure}[h]
 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability.pdf}
 	\caption{Maintainability difference for security and baseline refactorings}
	\label{fig:secvsreg}
\end{figure}

@@

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
affect open-source software maintainability?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.2cm}
